---
title: "A novel role for alkyladenine DNA glycosylase in regulating alkylation-induced ER stress"
subtitle: "Figure 1: Preparing expression data"
author: "L Milano, CF Charlier, R Andreguetti, A Aljohani, E Healing, MP Thomé, R Elliott, LD Samson, G Lenz, JAP Henriques, A Nohturfft and LB Meira"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "hide"
    number_sections: FALSE
    theme: "readable"
    highlight: "tango"
    fig_caption: TRUE
    css: "./source/styles.css"
---

```{js}
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
```

## Setup  

* Check whether required packages are installed  
* Clear memory and console  
* Load packages  
* Get current script name  
* Get project folder  
* Define function(s)  

```{r CHECK PACKAGES, results="hide"}
source("./source/check_installed_packages.source.R")
```


```{r CLEAN SLATE, results="hide"}
# Clear memory
rm(list=setdiff(ls(all=TRUE), c(".Random.seed")))
# Clear console
cat("\014")
```

```{r LOAD PACKAGES, results="hide"}
library(magrittr)
library(scales)
library(dplyr)
library(mouse430a2.db)
```


```{r SCRIPT INFO, class.output="txt_output"}
this.script <- rstudioapi::getActiveDocumentContext()$path %>% basename
stopifnot(this.script != "")
cat("Script:", this.script)
```

```{r PROJECT DIRECTORY, results="asis", include=FALSE}
cat("Project folder:", rstudioapi::getActiveProject())
```


```{r DT OPTIONS}
dt.options <- function(x, align="left") {
  align <- paste0("dt-", align)
  list(dom="t",
       columnDefs = list(list(className = align,
       targets = seq_len(ncol(x))-1)))
}
```


## Load data  
### Get Affymetrix CEL files  
```{r DATA FOLDERS}
folder.cel <- "~/_Data/Aag/Aag_KO_Liver_Project_2/microarray/microarray_input/cel_files_paper"
stopifnot(all(file.exists(folder.cel)))
```

```{r CEL DATA FOLDER, results="asis"}
details::details(paste0("<tt>", folder.cel, "</tt>"), lang=NULL, summary="Show data folder")
```

```{r GET NAMES OF CEL FILES, results="asis"}
file.paths <- list.files(path=folder.cel, pattern=".*\\.CEL", full.names=TRUE)
details::details(paste("<tt>", paste(sort(basename(file.paths)), collapse="<br\n>"), "</tt>", sep="\n"),
                 lang=NULL, summary="Show CEL filenames")
```

### Read CEL files  
```{r READ CEL FILES, class.output="txt_output"}
stopifnot(all(file.exists(file.paths)))
ab <- affy::ReadAffy(filenames=file.paths)
Biobase::sampleNames(ab) <- basename(file.paths)
cat("Size of ", class(ab), ": ", utils:::format.object_size(pryr::object_size(ab), "auto"), sep="") # 81.8 Mb
```

```{r PRINT AFFYBATCH INFO, warning=FALSE}
print(ab)
```

```{r REMOVE OBSOLETE VARIABLES}
rm(file.paths, folder.cel)
```


## Add phenodata to affybatch
### Define pData  
```{r}
pd <- data.frame(sample.name = Biobase::sampleNames(ab),
                 Genotype=NA,
                 stringsAsFactors = F)
pd$Genotype <- ifelse(grepl("WT", pd$sample.name, ignore.case=T), "wt", "ko")
stopifnot(all(complete.cases(pd)))
pd$Replicate <- stringr::str_extract(pd$sample.name, "[0-9].CEL$") %>% stringr::str_extract(., "^[0-9]")
pd$Treatment <- ifelse(grepl("Ctrl", pd$sample.name, ignore.case=T), "ctrl", "mms")
rownames(pd) <- pd$sample.name %>% sub("^AA[0-9]{3}_", "", .) %>% sub("^Mpg", "", .) %>% sub(".CEL$", "", .)
pd <- pd %>% dplyr::select(-sample.name)
pd %>% 
  tibble::rownames_to_column("Row Label") %>% 
  DT::datatable(., options = dt.options(.), rownames = F)
```

### Define varLabels  
```{r}
v.labels <- list(Genotype = "Mpg genotype",
                 Replicate = "1 to 3 arbitrary numbering",
                 Treatment = "Control or 6 hours MMS") %>% unlist
metaData <- data.frame(labelDescription=v.labels)
metaData %>% 
  tibble::rownames_to_column("Row Label") %>% 
  DT::datatable(., options = dt.options(.), rownames = F)
```

### Compose phenoData:  
```{r}
phenoData <- new("AnnotatedDataFrame",
                 data=pd, varMetadata=metaData)
Biobase::phenoData(ab) <- phenoData
# class(Biobase::phenoData(ab)) # "AnnotatedDataFrame"
Biobase::phenoData(ab)
```

```{r}
Biobase::protocolData(ab) <- Biobase::phenoData(ab)
```


```{r}
rm(v.labels, metaData, phenoData)
```


### Checking: print pData
```{r}
Biobase::pData(ab) %>% 
  tibble::rownames_to_column("Row label") %>% 
  DT::datatable(., options = dt.options(.), rownames = F)
```
### Save affybatch  
```{r SAVE AFFYBATCH, class.output="txt_output"}
if (!file.exists("data")) {
  dir.create("data")
}
out.file.affybatch <- paste0("./data/", this.script, ".affybatch.RDS")
saveRDS(ab, file=out.file.affybatch, compress=T)
cat("Saved:", out.file.affybatch)
rm(out.file.affybatch)
```

## Generate expression set  
Following instructions in affy vignette (see Info/affy_vignette_2017_04_24.pdf). *"The function expresso performs the steps background correction, normalization, probe specic correction, and summary value computation."* ([Gautier et al., 2004](https://www.ncbi.nlm.nih.gov/pubmed/14960456)).    

```{r GENERATE EXPRESSION SET, results="hide"}
eset <- affy::expresso(ab,
                       normalize.method="qspline",
                       bgcorrect.method="rma",
                       pmcorrect.method="pmonly",
                       summary.method="liwong",
                       verbose = FALSE)
```

### Save eset  
```{r SAVING ESET, class.output="txt_output"}
out.file.eset <- paste0("./data/", this.script, ".eset.RDS")  
saveRDS(eset, file=out.file.eset, compress=TRUE)
cat("Saved:", out.file.eset)
rm(out.file.eset)
```

### Boxplot of log2 transformed data
```{r BOXPLOT ESET, message=FALSE, warning=FALSE}
op <- par()
par(mar=c(7,4,0,2)) # default is (5,4,4,2)
eset %>% Biobase::exprs(.) %>% log2(.) %>%
  boxplot(., las=2, col=rep(RColorBrewer::brewer.pal(4,"Set2"), each=3))
par(op)
rm(op)
```

```{r results="hide"}
expr.df <- eset %>% Biobase::exprs(.) %>% log2(.) %>% as.data.frame %>% tibble::rownames_to_column("ID_REF")
```

```{r results="hide"}
# Print first 6 rows:
head(expr.df)
```


### Save expression table with all replicates:  
```{r SAVING MATRIX, class.output="txt_output"}
out.file.expr.all <- paste0("./data/", this.script, ".expr_all.txt")
write.table(expr.df, file=out.file.expr.all, sep="\t", col.names = T, row.names = F)
cat("Saved:", out.file.expr.all)
rm(out.file.expr.all)
```

## Calculate average expression values  
### Print first six rows  
```{r}
expr.df.means <- expr.df %>% 
  dplyr::rowwise(ID_REF) %>% 
  dplyr::mutate(Wt_Ctrl = mean(c_across(starts_with("WT_Ctrl"))),
         Wt_MMS = mean(c_across(starts_with("WT_MMS"))),
         Ko_Ctrl = mean(c_across(starts_with("KO_Ctrl"))),
         Ko_MMS = mean(c_across(starts_with("KO_MMS")))) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(-starts_with("WT_Ctrl", ignore.case=F)) %>%
  dplyr::select(-starts_with("WT_MMS", ignore.case=F)) %>% 
  dplyr::select(-starts_with("KO_Ctrl", ignore.case=F)) %>% 
  dplyr::select(-starts_with("KO_MMS", ignore.case=F))
head(expr.df.means) %>% 
  DT::datatable(., options = dt.options(.), rownames = F) %>% 
  DT::formatRound(setdiff(names(expr.df.means), "ID_REF"), digits=3)
```


### Add gene annotations  
```{r results="hide"}
anno <- AnnotationDbi::select(mouse430a2.db, row.names(eset), c("SYMBOL", "GENENAME"), "PROBEID")
head(anno)
```

```{r}
expr.df.means.anno <- merge(x=anno, y=expr.df.means,
                              by.x="PROBEID",
                              by.y="ID_REF", all=TRUE)
rm(anno)
```

### Print first six rows  
```{r}
head(expr.df.means.anno)
```
<div class="comments">
*Note that probe IDs and gene symbols are no longer unique after adding gene annotations!*
</div>

```{r class.output="txt_output"}
cat("Number of duplicated probe IDs: ", comma(sum(duplicated(expr.df.means.anno$PROBEID))), "\n",
    "Number of duplicated gene symbols: ", comma(sum(duplicated(expr.df.means.anno$SYMBOL))), "\n",
    "Number of missing probe IDs: ", comma(sum(is.na(expr.df.means.anno$PROBEID))), "\n",
    "Number of missing gene symbols: ", comma(sum(is.na(expr.df.means.anno$SYMBOL))),
    sep="")
```


### Save average expression data  
```{r class.output="txt_output"}
out.file.expr.all <- paste0("./data/", this.script, ".expr_mean.txt")
write.table(expr.df.means.anno, file=out.file.expr.all, sep="\t", row.names=FALSE, col.names=TRUE)
cat("Saved:", out.file.expr.all)
rm(out.file.expr.all)
```


## Session info  

<button class="button" onclick="myFunction('DIV_1')">Show/hide session info</button>
<div id="DIV_1" class="div_default_hide">

```{r SESSION INFO DATE, results="asis"}
cat("Date:", format(Sys.time(), "%a %d-%b-%Y %H:%M:%S"), "<br>\n")
```

```{r print_session_info, R.options=list(width=70)}
devtools::session_info()
```
</div>

```{js}
var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
divsToHide[i].style.display = 'none';
}
```
