---
title: "A novel role for alkyladenine DNA glycosylase in regulating alkylation-induced ER stress"
subtitle: "Supplemtal Table 1"
author: "L Milano, CF Charlier, R Andreguetti, A Aljohani, E Healing, MP Thomé, R Elliott, LD Samson, G Lenz, JAP Henriques, A Nohturfft and LB Meira"
output:
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "hide"
    number_sections: FALSE
    theme: "readable"
    highlight: "tango"
    fig_caption: TRUE
    css: "./source/styles.css"
---

```{js}
function myFunction(id) {
  var x = document.getElementById(id);
  if (x.style.display === 'none') {
    x.style.display = 'block';
  } else {
    x.style.display = 'none';
  }
}
```


## Setup  

* Clear memory and console  
* Load packages  
* Get current script name  
* Get project folder  
* Define function(s)  


```{r CLEAN SLATE, results="hide"}
# Clear memory
rm(list=setdiff(ls(all=TRUE), c(".Random.seed")))
# Clear console
cat("\014")
```

```{r LOAD PACKAGES, results="hide"}
library(magrittr)
library(dplyr)
library(AnnotationDbi)
library(mouse430a2.db)
```


```{r, SCRIPT INFO}
this.script <- rstudioapi::getActiveDocumentContext()$path %>% basename
stopifnot(this.script != "")
cat("Script:", this.script)
```

```{r PROJECT DIRECTORY, results="asis", results="hide"}
cat("Project folder:", rstudioapi::getActiveProject())
```


```{r DT OPTIONS}
dt.options <- function(x, align="left") {
  align <- paste0("dt-", align)
  dom.value <- ifelse(nrow(x) > 5, "tip", "t")
  list(dom=dom.value,
       columnDefs = list(list(className = align,
       pageLength = 10,
       targets = seq_len(ncol(x))-1)))
}
```

## Input data  
Script expects that input data have been generated in previous script.  

### Load expression set  
```{r LOAD ESET}
# list.files("data")
eset.file <- "./data/01_Raw_Data_Processing.Rmd.eset.RDS"
stopifnot(file.exists(eset.file))
eset <- readRDS(eset.file)
cat("Read eset from:", eset.file)
rm(eset.file)
```

```{r SIZE ESET, results="asis"}
cat("Class: <b>", class(eset), "</b><br>\n",
    "Size: <b>",
    utils:::format.object_size(pryr::object_size(eset), "auto"),
    "</b>\n", sep="") # 6.3 Mb
```

### Load gene annotations  
```{r results="hide", message=FALSE, warning=FALSE}
anno <- AnnotationDbi::select(mouse430a2.db, row.names(eset), c("SYMBOL", "GENENAME"), "PROBEID")
head(anno)
```

## Limma analysis  
### Sample table  
```{r SAMPLE TABLE}
sample.table <- Biobase::pData(eset)
sample.table
```

### Experimental design  
```{r EXPERIMENTAL DESIGN, results="hide"}
TS <- paste(sample.table$Genotype, sample.table$Treatment, sep=".")
TS <- factor(TS, levels=c("wt.ctrl","wt.mms","ko.ctrl","ko.mms"))
TS
```

<button class="button" onclick="myFunction('DIV_1')">Show/hide model matrix</button>
<div id="DIV_1" class="div_default_hide">
```{r MODEL MATRIX, results="asis"}
design <- model.matrix(~0+TS)
colnames(design) <- levels(TS)
design
```
</div>

### Fitting linear model  
```{r LINEAR MODEL}
fit <- Biobase::exprs(eset) %>% 
  log2(.) %>% 
  limma::lmFit(., design)
cat("Object class:\n")
class(fit)
```


### Contrast matrix  
```{r}
cont.matrix <- limma::makeContrasts(
     ko.ctrl.vs.wt.ctrl = ko.ctrl - wt.ctrl,
     ko.mms.vs.wt.mms   = ko.mms - wt.mms,
     wt.mms.vs.wt.ctrl  = wt.mms - wt.ctrl,
     ko.mms.vs.ko.ctrl  = ko.mms - ko.ctrl,
     levels=design)
cont.matrix
```


### Statistics for Differential Expression  

```{r FIT2, results="asis"}
fit2 <- limma::contrasts.fit(fit, cont.matrix)
fit2 <- limma::eBayes(fit2)
details::details(fit2, lang=NULL, summary="Show results of limma::eBayes")
```

### Save statistics  
```{r}
out.file.fit <- paste0("./data/", this.script, ".fit.RDS")
saveRDS(object=fit2, file=out.file.fit, compress = TRUE)
cat("Saved:", out.file.fit)
rm(out.file.fit)
```


## Number of differentially expressed genes: 'global' versus 'separate' methods  
The `limma::decideTests` function identifies "which genes are significantly differentially expressed for each contrast from a fit object containing p-values and test statistics."  

"The setting method="separate" is equivalent to using topTable separately for each coefficient in the linear model fit, and will identify the same probes as significantly differentially expressed if adjust.method is the same. method="global" will treat the entire matrix of t-statistics as a single vector of unrelated tests." [...] 

### Regulation parameters (cutoffs log2FC & p.value):  
```{r P_CUTOFF, results="asis"}
p.cutoff <- 0.05
cat("Maximum p value:<b>", p.cutoff, "</b>")
```

```{r LOGFC_CUTOFF, results="asis"}
fc.min <- 1.75
lfc.min <- log2(fc.min)
cat("Minimum log2FC:<b>", paste0("log2(", fc.min, ") = ", round(lfc.min, 3)), "</b>")
```

### DecideTests  
```{r}
test.results <- limma::decideTests(fit2, p.value = p.cutoff, lfc=lfc.min,
                                  method = "separate", adjust.method="fdr")
```


```{r}
summary(test.results) %>% unclass %>% as.data.frame %>% .[-2,] %>% 
  rbind(., colSums(.)) %>% magrittr::set_rownames(c("Down", "Up", "Total"))
```




## Get differentially regulated probes  
```{r}
mx <- unclass(test.results)
test.results.probes <- vector("list", ncol(mx)) %>% 
  set_names(colnames(mx))

get.probes <- function(mtx, colm, where) {
  where <- tolower(where)
  stopifnot(colm %in% colnames(mtx))
  stopifnot(where %in% c("up", "down"))
  where2 <- ifelse(where == "up", 1, -1)
  a <- mtx[,colm]
  b <- a[a == where2]
  c <- names(b)
  d <- sort(c)
  d
}

test.results.probes <- lapply(seq_len(ncol(mx)), function(i) {
  probes.down <- get.probes(mx, colnames(mx)[i], "down")
  probes.up <- get.probes(mx, colnames(mx)[i], "up")
  probes.total <- sort(union(probes.up, probes.down))
  list(down=probes.down, up=probes.up, total=probes.total)
}) %>% 
  set_names(colnames(mx))

str(test.results.probes)
```


### Save probe lists  
```{r}
out.file.probe.lists <- paste0("./data/", this.script, ".DEprobes.RDS")
saveRDS(object=test.results.probes, file=out.file.probe.lists)
cat("Saved:", out.file.probe.lists)
rm(out.file.probe.lists)
```


## Get differentially regulated gene symbols  
```{r}
get.gene.symbols <- function(probe.ids) {
  stopifnot(all(probe.ids %in% anno$PROBEID))
  gs <- anno %>% 
    filter(PROBEID %in% probe.ids) %>% 
    tidyr::drop_na(.) %>% 
    .$SYMBOL %>% unique %>% sort
  gs
}
```


```{r}
test.results.genes <- lapply(test.results.probes, function(probs) {
  lapply(probs, get.gene.symbols)
})
str(test.results.genes)
```


```{r}
lapply(test.results.genes, function(gs) {
  sapply(gs, length) %>% as.matrix
}) %>% do.call(cbind, .) %>% 
  set_colnames(names(test.results.genes))
```

#### Save gene lists  
```{r}
out.file.gene.lists <- paste0("./data/", this.script, ".DEgenes.RDS")
saveRDS(object=test.results.genes, file=out.file.gene.lists)
cat("Saved:", out.file.gene.lists)
rm(out.file.gene.lists)
```


## Toptables  
```
NOTE: After annotating the toptables with gene symbols, the number of differentially regulated items increases because some probeset IDs map to several gene symbols.
```

### Annotate probe sets with gene symbol and gene title:  
```{r}
# 'anno' defined early in script
add.anno <- function(dafra) {
  first.three.columns <- c("PROBEID", "SYMBOL", "GENENAME")
  remaining.columns <- setdiff(names(dafra), first.three.columns)
  cols <- c(first.three.columns, remaining.columns)
  dafra <- dafra %>% tibble::rownames_to_column("PROBEID") %>% 
    dplyr::left_join(anno, by="PROBEID") %>% 
    dplyr::select(cols) %>% 
    dplyr::arrange(P.Value) %>% 
    magrittr::set_rownames(NULL)
}
```

### Generate toptables  
```{r}
# coefs <- c("ko.ctrl.vs.wt.ctrl", "ko.mms.vs.wt.mms")
coefs <- colnames(cont.matrix)
tt <- lapply(coefs, function(coeff) {
  limma::topTable(fit=fit2, number=Inf, coef=coeff) %>% 
    add.anno(.) %>%
    dplyr::arrange(desc(logFC))
}) %>% 
  magrittr::set_names(coefs)
```

```
NOTE: Many probe IDs map to several gene symbols:  
```

```{r results="asis"}
cat("Rows in eset:", scales::comma(nrow(eset)))
```

```{r}
sapply(tt, nrow) %>% scales::comma(.) %>%
  as.data.frame(., stringsAsFactors=FALSE) %>% 
  tibble::rownames_to_column(.) %>% 
  set_colnames(c("Contrast", "Rows"))
```

### Save toptables in Excel file  
```{r results="asis"}
xlsx.file <- "Supplemental_Table_1.xlsx"
wb <- openxlsx::createWorkbook()
openxlsx::modifyBaseFont(wb, fontSize = 12, fontName = "Arial")
tab.colors <- c("paleturquoise4", "darkseagreen", "palevioletred", "bisque3")
for (i in seq_along(tt)) {
  sheet.name <- names(tt)[i]
  openxlsx::addWorksheet(wb=wb, sheetName=sheet.name, gridLines = FALSE, tabColour=tab.colors[i])
  openxlsx::setColWidths(wb=wb, sheet=sheet.name, cols = 1, widths = "auto")
  openxlsx::writeDataTable(wb = wb, sheet=sheet.name, x = tt[[i]],
                           colNames = TRUE)
}
openxlsx::saveWorkbook(wb, xlsx.file, overwrite = TRUE)
cat("Saved:", basename(xlsx.file))
rm(xlsx.file, wb, tab.colors, i, sheet.name)
```


## Session info  

<button class="button" onclick="myFunction('DIV_2')">Show/hide session info</button>
<div id="DIV_2" class="div_default_hide">

```{r SESSION INFO DATE, results="asis"}
cat("Date:", format(Sys.time(), "%a %d-%b-%Y %H:%M:%S"), "<br>\n")
```

```{r print_session_info, R.options=list(width=70)}
devtools::session_info()
```
</div>
  
```{js}
var divsToHide = document.getElementsByClassName("div_default_hide");
for(var i = 0; i < divsToHide.length; i++)
{
  divsToHide[i].style.display = 'none';
}
```
